Factors Influencing Abundance & Virus Isolation in Mosquitoes

### 2024-04-30 R/analysis-code/analysis.qmd
edits to readme and mauscript

>>>>

# Extract the fit using the appropriate tidymodels function
fitted_model <- extract_fit_engine(final_fit_continuous)

# Check if the extracted model is the correct type
if(inherits(fitted_model, "ranger")) {
    # If it is the correct type, extract feature importance
    importance_matrix <- fitted_model$variable.importance
    feature_importances <- tibble(
        Feature = names(importance_matrix),
        Importance = importance_matrix
    ) %>%
    arrange(desc(Importance))

    # Print the feature importances
    print(feature_importances)

    # Plot the feature importances
    importance_plot <- ggplot(feature_importances, aes(x = reorder(Feature, Importance), y = Importance)) +
        geom_col(fill = "steelblue") +
        coord_flip() +
        labs(title = "Feature Importance from Random Forest", x = "Importance", y = "Features") +
        theme_minimal()

    # Save the plot
    ggsave("feature_importance_rf.pdf", plot = importance_plot, width = 8, height = 6)
} else {
    cat("The extracted object is not a 'ranger' Random Forest model. Please check the model setup and extraction.\n")
}
print(importance_plot)

### 2024-04-30 products/manuscript/Manuscript.qmd
Debug and clean
debugged code and updated README files
edits to readme and mauscript

# Problem statement

Which environmental,ecological, and temporal factors influence the
transmission cycles for EEE in southeastern Massachusetts and which
models best describe the relationship between these factors?

@@

Bristol County, Massachusetts covers an area of 691 square miles with a
population density of 962 people per square mile. It borders Rhode
Island to the west, Plymouth County, MA to the east, and Norfolk County,
MA to the north. Naushon Island and Martha's Vineyard are across
Buzzard's Bay to the Southeast.

Atlantic White Cedar (AWC) swamps span over 4,300 acres within
Massachusetts, primarily concentrated in Plymouth and Bristol counties.
These swamps serve as critical breeding habitats for *Culiseta melanura*
(MEL)
[@commonwealthofmassachusettsdivisionoffisheries-wildlifeNaturalCommunityFact2007],
which is recognized as the primary enzootic vector for Eastern Equine
Encephalitis (EEE). AWC trees have large root systems that form "crypts"
and hold water throughout the year. These crypts serve as a main habitat
for MEL larval growth and provide shelter during winter months.

@@

**Which environmental,ecological, and temporal factors influence the
transmission cycles for EEE in southeastern Massachusetts and which
models best describe the relationship between these factors?** From
prior research we know that the mosquito lifecyle is influenced by a
number of factors. Temperature plays a major role in the mosquito life
cyle virus transmission dynamics. As temperature increases, the rate of
larval development also increases and the time between when a mosquito
ingests virus in a bloodmeal and when it can transmit it decreases. The
result is that the mosquito population can grow at faster rate and they
will have more time to transmit the virus. However, MEL larvae develop
in well-insulated crypts that are less sensitive to ambient air
temperature changes, so the effect of temperature changes on MEL
development speed may be less pronounced than other vector species that
develop in open, standing water, like ditches and containers.
Precipitation increases the amount of standing water and increases the
chances of larval survival during the mosquito season and also during
the winter months for those species that overwinter in the larval stage.
For EEE, the main enzootic vector, MEL, overwinters as larvae.

@@

The mosquito surveillance dataset encompasses counts by species from all
mosquito traps deployed in Bristol County, MA, from 2007 to 2023. The
raw dataset includes over 45,000 records and includes temporal and
spatial variable. The BCMCP, perform mosquito surveillance from June
through October annually.

During this interval, CDC light traps baited with carbon dioxide are
placed at various sentinel sites each week. Designed to attract
mosquitoes in search of blood meals, light traps serve as a reliable
measure of the actively feeding mosquito population, thereby
establishing them as the standard method for EEE surveillance. Some of
the sentinel site have been used for over 50 years.

Additionally, the county employs gravid traps, which are specifically
aimed at attracting and capturing gravid *Culex* mosquitoes looking for
egg-laying sites. While these traps are primarily used for West Nile
Virus (WNV) surveillance, they hold significant value for EEE monitoring
as well. Given the role of *Culex* species as bridge vectors for EEE,
their capture in gravid traps provides critical insight into the risk of
virus transmission.

The traps are installed in the afternoon and retrieved the following
morning. After the traps are collected, the next are stored in a freezer
to preserve and kill the mosquitoes. The mosquitoes are sorted by
species using a dissecting scope. Counts for each species are recorded
in the database and a portion of the vector species are driven to the
Massachusetts State Laboratory in Boston. The lab tests each mosquito
pool (or batch) for WNV and EEEv using PCR. Results are shared with
BCMCP on the same day of testing. Virus isolation results are recorded
in the collection records.

@@

One indicator of the level of virus activity present in the mosquito
population can be calculated from the information contained in the
mosquito dataset. The minimum infection rate (MIR) can be calculated at
various temporal and spatial scales from trap site over one day to all
sites over an entire season. MIR is calculated as: $$
MIR = \left( \frac{\text{Number of positive mosquito pools}}{\text{Total number of mosquitoes tested}} \right) \times 1000
$$ MIR provides an estimate of the prevalence of a specific pathogen
within a mosquito population. While it does not directly measure the
risk to humans, a higher MIR in mosquitoes that commonly bite humans can
indicate a higher risk of transmission. A study pubulihsed in 2007
indicated that in years with human cases in Massachusetts between 1979
and 2004, MIR was ≥1.0 by week 31 [@hachiyaHumanEasternEquine2007]. MIR
allows for the comparison of infection rates across different
geographical areas or time periods. This is crucial for identifying
areas or times of higher risk and for evaluating the effectiveness of
control measures.

@@

The hourly summaries were used to calculate weather variables, such as
the daily maximum temperature and precipitation. While daily summaries
provide this information will less processing, the hourly summaries
provide more granular data that can be used to calculate additional
variables, such as the number of hours above a certain temperature
threshold or degree-days. Degree-days is a measure of the amount of heat
accumulated over a period of time. It is calculated by summing the
difference between the average daily temperature and a base temperature.
Degree-days are used to estimate the rate of development of insects and
other organisms that are sensitive to temperature. For mosquitoes,
degree-days can be used to estimate the rate of larval development and
the time between when a mosquito ingests a virus and when it can
transmit it [@guStatisticalEstimationDegree2006]. To simplify this
variable, the number of hours the temperature was ≥50°F and ≤86°F. These
thresholds were chosen based on the known temperature range for mosquito
development and virus transmission.

@@

## Data aquisition

@@

The Massachusetts Department of Public Health requires that all mosquito
control districts submit their collection data in a specific format.
This format divides each trap event into separate rows based on mosquito
species and whether they were submitted for virus testing. However, this
format is not useful for certain calculations where a zero count for a
species is not explicitly recorded. For example, even when sorted by
species to calculate the average count per trap, the average could be
inflated since the records will not include trap events with a zero
count of that species. In order to correct this, the data was pivoted to
create a column for every unique "species code" then the "pool size" for
each species code is summed for each trap event. A trap event is when
"town", "date of collection", "trap type" are all the same. For species
codes not included in the orginal table for a trap event, the field will
be assigned a zero value. The resulting table will have one row for
every trap event.

@@

Symptomatic EEE, and to a lesser extent WNV, human cases are rare. On
average, the United States records only about 9 cases of EEE annually.
Notably, Massachusetts reports more cases than any other state except
Florida, highlighting regional disparities in disease occurrence.
However, the incidence rate for neuroinvasive EEE cases in Bristol
County, Massachusetts, is estimated to be only around 0.06 per 100,000
people
[@centersfordiseasecontrolandpreventionPreventionEasternEquine2022]. The
low incidence of human cases poses significant challenges for using
traditional statistical models, making it difficult to accurately
estimate human transmission risk. This difficulty is largely due to the
sparse nature of the data and the variability in transmission patterns,
which complicates efforts to predict outbreaks and necessitates
alternative modeling approaches.

MIR will be the main outcome of interest for this project. MIR values
will be calculated for each week during the surveillance season over the
entire county. To reduce bias, only mosquitoes captured using carbon
dioxide-baited CDC light traps will be used to to calculate MIR. These
traps are the standard EEE surveillance and are used to monitor the
actively feeding mosquito population. Additionally, these trap types
have been used in this area for over 40 years, many at the same sites
currently in use.

@@

In the first stage of the analysis, the GAM was utilized to address the
high incidence of zeros in the dataset—an indicator of overdispersion.
This zero-inflated model allowed us to separately model the presence
versus absence of the outcome variable, thereby appropriately accounting
for the excessive zeros.

Subsequently, the second stage of the GAM focused on modeling the
continuous part of the MIR data, relating it to the suite of potential
predictors. This step incorporated smooth functions of the predictors,
permitting the data to inform the shape of the relationship without
imposing a predetermined form, as would be the case in a traditional
linear model.

While the two-stage GAM appeared to be a good fit, there were too few
observations to provide a robust estimate of the relationship between
the predictors and MIR. The original dataset contained over 40,000
observations, but after reformatting the data to fit a rectangular shape
needed for analysis, there were less than 400 observations in the final
set. After the zeros were removed for part 2 of the GAM, the sample
number of observations left was less than 100. Because of this, it was
decided to use a single GAM with a dataset that included zero and
non-zero observations. Additionally, all trap types were included in the
analysis, not just the carbon dioxide-baited CDC light traps. While some
precision was lost when the one part model was adopted, the gain in
observations was deemed more valuable.

### Schematic of workflow

*Schematic will be refined in later steps. The doc will also be
reformatted to move the schematic to an appendix* *The image should
appear upon rendering*

```{mermaid}
flowchart TB
    id1["Questions"] --> id2["Data Collection"]
    id2 --> id3["Data Loading and Cleaning"]
    id3 --> id4["Data Cleaning"]
    id4 --> id5["Data Transformation"]
    id5 --> id6["Data Exploration & Visualization"]
    id6 --> id7["Data Modeling"]
    id7 --> id8["Evaluation"]

    id8 -.-> id1
    id8 --> id9[Interpretation]
    id9 --> id10[Communication]

    classDef mainProcess fill:#fff,stroke:#333,stroke-width:2px;

    class id1,id2,id3,id4,id5,id6,id7,id8,id9,id10 mainProcess
```

@@

There was one other significant result. When MIR is set as the
dependent, continuous variable in a linear regression and the average
maximum temperature for the previous week is used as the predictor, the
model suggests there is a significant inverse relationship. As average
temperature increases, the MIR decreases (model_mir_temp.R). This
relationship is counterintutive, especially given the result of the
logistic regression and requires more examination.

@@

@@

Analysis began with a Generalized Additive Model (GAM) with predictors
that were selected based on the exploratory single variable GAM. The
variables were chosen if their p=values were \<0.05. Eight variables met
that condition including: the number of hours below 50°F the week before
the mosquito collection week (Hrs_Below_50_lag), the number of hours
below 50°F the week of collection (Hrs_Below_50_current), the drought
severity index for East Coast FEMA regions (DSCI_fema), the average
hourly relative humidity the week of collection
(Avg_HourlyRelativeHumidity_current), the average hourly windspeed the
week of collection (Avg_HourlyWindSpeed_current), the average number of
daylight hours per day (average_daylight_hours), the average number of
MEL captured by trap (Average_MEL), and the proportion of MEL captured
out of the total number of mosquitoes trapped (MEL_Prop).

@@

![](images/rmse_table-01.png)

@@

The null model RMSE was 2.40. All models showed improvement over the
null model on both the training and testing datasets. The GAM shows
significant improvement over the null model for the training set, but
only slightly improves over the null for the testing set. This suggests
that beyond capturing the mean response, this GAM is not well
generalized. The substantial improvement in the training set RSME for
GBM and Random Forest coupled with less improvement over the null for
the testing set, suggest a tendency for overfitting. These models may be
capturing the training data's patterns that do not gerneralize well to
unseen data. The the gap between the training and testing RMSE for the
LASSO and GAMS models suggest there is less overfitting and are slightly
better at generalizing than the others. However, given the non-linear
relationships, the LASSO model is likely not appropriate for this
analysis.

@@

Likely the biggest limitation \## Conclusions While all models
demonstrated improvement over null model, none of them were robust
enough to be used for reliable prediction. In the next phases of this
project, the data will be expanded to include more years and cover a
larger portion of Massachusetts. In additon to increasing the the number
of observations, other sources of data will be incorporated. Including
virus activity from surrounding regions and areas from which birds
migrate to Massachusetts.

>>>>

# Research question

Which environmental,ecological, and temporal factors influence the
transmission cycles for EEE in southeastern Massachusetts, and which
models best describe the relationship between these factors?

@@

Bristol County, Massachusetts covers 691 square miles with a population
density of 962 people per square mile. It borders Rhode Island to the
west, Plymouth County, MA to the east, and Norfolk County, MA to the
north. Naushon Island and Martha's Vineyard are across Buzzard's Bay to
the Southeast.

Atlantic White Cedar (AWC) swamps cover more than 4,300 acres across
Massachusetts, with the largest concentrations found in Plymouth and
Bristol counties. These swamps are crucial breeding grounds for Culiseta
melanura (MEL), which is identified as the primary enzootic vector for
Eastern Equine Encephalitis (EEE)
[@commonwealthofmassachusettsdivisionoffisheries-wildlifeNaturalCommunityFact2007].
Atlantic White Cedar (AWC) trees possess extensive root systems that
create "crypts" capable of retaining water year-round. These crypts are
crucial habitats for the larval growth of Culiseta melanura (MEL) and
also provide shelter during the winter months.

@@

**Which environmental,ecological, and temporal factors influence the
transmission cycles for EEE in southeastern Massachusetts, and which
models best describe the relationship between these factors?** From
prior research we know that the mosquito lifecyle is influenced by a
number of factors. Temperature plays a major role in the mosquito life
cyle virus transmission dynamics. As temperature increases, the rate of
larval development also increases and the time between when a mosquito
ingests virus in a bloodmeal and when it can transmit it decreases. The
result is that the mosquito population can grow at faster rate and they
will have more time to transmit the virus. However, MEL larvae develop
in well-insulated crypts that are less sensitive to ambient air
temperature changes, so the effect of temperature changes on MEL
development speed may be less pronounced than other vector species that
develop in open, standing water, like ditches and containers.
Precipitation increases the amount of standing water and increases the
chances of larval survival during the mosquito season and also during
the winter months for those species that overwinter in the larval stage.
For EEE, the main enzootic vector, MEL, overwinters as larvae.

@@

The mosquito surveillance dataset encompasses counts by species from all
mosquito traps deployed in Bristol County, MA, from 2007 to 2023. The
raw dataset includes over 45,000 records and includes temporal and
spatial variables. The BCMCP, perform mosquito surveillance from June
through October annually.

During this period, CDC light traps, baited with carbon dioxide, are
strategically placed at various sentinel sites weekly. These traps are
designed to attract mosquitoes seeking blood meals and are recognized as
a reliable method for monitoring the actively feeding mosquito
population. Consequently, they are established as the standard approach
for Eastern Equine Encephalitis (EEE) surveillance. Notably, some of
these sentinel sites have been in continuous use for over 50 years.

Additionally, the county deploys gravid traps, which are specifically
aimed at attracting and capturing gravid *Culex* mosquitoes looking for
egg-laying sites. While these traps are primarily used for West Nile
Virus (WNV) surveillance, they hold significant value for EEE monitoring
as well. Given the role of *Culex* species as bridge vectors for EEE,
their capture in gravid traps provides critical insight into the risk of
virus transmission.

The traps are set up in the afternoon and collected the following
morning. Once retrieved, the captured mosquitoes are immediately stored
on dry ice and then in a freezer. This process kills the mosquito and
preserves them. The mosquitoes are sorted by species under a dissecting
scope. Counts for each species are recorded in the database and a
portion of the vector species are driven to the Massachusetts State
Laboratory in Boston. The lab tests each mosquito pool (or batch) for
WNV and EEEv using PCR. Results are shared with BCMCP on the same day of
testing. Virus isolation results are recorded in the collection records.

@@

In our investigation into the environmental, ecological, and temporal
factors that impact the transmission cycles of EEE in southeastern
Massachusetts, the Mosquito Infection Rate (MIR) emerges as a pivotal
indicator. MIR provides a quantifiable measure of virus prevalence
within mosquito populations, making it an essential metric for modeling
the dynamics of disease spread. By analyzing changes in MIR in response
to varying environmental conditions and mosquito behaviors, we can
enhance our understanding of the key drivers behind EEE transmission.
This understanding, in turn, guides the development of predictive models
that can accurately forecast periods of heightened transmission risk
based on observed or anticipated changes in the identified factors. The
minimum infection rate (MIR) can be calculated at various temporal and
spatial scales from trap site over one day to all sites over an entire
season. MIR is calculated as: $$
MIR = \left( \frac{\text{Number of positive mosquito pools}}{\text{Total number of mosquitoes tested}} \right) \times 1000
$$ While MIR does not directly measure the risk to humans, a higher MIR
in mosquitoes that commonly bite humans can indicate a higher risk of
transmission. A study pubulihsed in 2007 indicated that in years with
human cases in Massachusetts between 1979 and 2004, the MIR was ≥1.0 by
week 31 (mid-late July) [@hachiyaHumanEasternEquine2007]. MIR allows for
the comparison of infection rates across different geographical areas or
time periods. This is crucial for identifying areas or times of higher
risk and for evaluating the effectiveness of control measures.

@@

The hourly summaries were used to calculate weather variables, such as
the daily maximum temperature and precipitation. While daily summaries
provide this information will less processing, the hourly summaries
provide more granular data that can be used to calculate additional
variables, such as the number of hours above a certain temperature
threshold. or degree-days. Degree-days is a measure of the amount of
heat accumulated over a period of time. It is calculated by summing the
difference between the average daily temperature and a base temperature.
Degree-days are used to estimate the rate of development of insects and
other organisms that are sensitive to temperature. For mosquitoes,
degree-days can be used to estimate the rate of larval development and
the time between when a mosquito ingests a virus and when it can
transmit it [@guStatisticalEstimationDegree2006]. To simplify this
variable, the number of hours the temperature was ≥50°F and ≤86°F. These
thresholds were chosen based on the known temperature range for mosquito
development and virus transmission.

@@

@@

The Massachusetts Department of Public Health mandates a specific format
for mosquito control districts to submit their collection data. In this
format, each trap event is divided into separate rows based on mosquito
species and whether they were submitted for virus testing. However, this
format poses limitations for certain calculations where a zero count for
a species is not explicitly recorded. For instance, when sorting data by
species to calculate the average count per trap, the absence of records
with a zero count could inflate the calculated average.

To address this issue, the data was transformed by pivoting it to create
a column for every unique "species code." Then, the "pool size" for each
species code was summed for every trap event, defined as instances where
"town," "date of collection," and "trap type" are identical. For species
codes not originally included in the table for a trap event, the
corresponding field was assigned a zero value. As a result, the
resulting table now contains one row for each trap event.

@@

Symptomatic EEE human cases are rare. On average, the United States
records only about 9 cases of EEE annually. Notably, Massachusetts
reports more cases than any other state except Florida, highlighting
regional disparities in disease occurrence. However, the incidence rate
for neuroinvasive EEE cases in Bristol County, Massachusetts, is
estimated to be only around 0.06 per 100,000 people
[@centersfordiseasecontrolandpreventionPreventionEasternEquine2022]. The
low incidence of human cases poses significant challenges for using
traditional statistical models, making it difficult to accurately
estimate human transmission risk. This difficulty is largely due to the
sparse nature of the data and the variability in transmission patterns,
which complicates efforts to predict outbreaks and necessitates
alternative modeling approaches.

MIR will be the main outcome of interest for this project. MIR values
will be calculated for each week during the surveillance season over the
entire county.

@@

In the first stage of the analysis, the GAM was utilized to address the
high incidence of zeros in the dataset. This zero-inflated model allowed
us to separately model the presence versus absence of the outcome
variable, thereby appropriately accounting for the excessive zeros.

The second stage of the GAM focused on modeling the continuous part of
the MIR data, relating it to the suite of potential predictors. This
step incorporated smooth functions of the predictors, permitting the
data to inform the shape of the relationship without imposing a
predetermined form, as would be the case in a traditional linear model.

While the two-stage Generalized Additive Model (GAM) initially seemed to
be a suitable approach, there were insufficient observations to yield a
robust estimate of the relationship between the predictors and Minimum
Infection Rate (MIR). Despite starting with a dataset containing over
40,000 observations, the transformation of the data into a rectangular
shape required for analysis resulted in fewer than 400 observations in
the final dataset. After removing the zeros for part 2 of the GAM, the
number of observations decreased to less than 100. Consequently, a
decision was made to utilize a single GAM with a dataset containing both
zero and non-zero observations. Although adopting a single-part model
led to some loss of precision, the increase in the number of
observations was deemed more advantageous.

@@

There was one other significant result. When MIR is set as the
dependent, continuous variable in a linear regression and the average
maximum temperature for the previous week is used as the predictor, the
model suggests there is a significant inverse relationship. As average
temperature increases the week prior to collection, the MIR decreases
(model_mir_temp.R). This relationship is counterintutive, especially
given the result of the logistic regression and requires more
examination.

@@

All detailed model analyses are available in the analysis.qmd file,
providing a transparent account of the methodologies employed and the
statistical rigor behind the conclusions drawn.

@@

Analysis began with a Generalized Additive Model (GAM) with predictors
that were selected based on the exploratory single variable GAM. The
variables were chosen if their p=values were \<0.05. Eight variables met
that condition including: the number of hours below 50°F the week before
the mosquito collection week (Hrs_Below_50_lag), the number of hours
below 50°F the week of collection (Hrs_Below_50_current), the drought
severity index for East Coast FEMA regions (DSCI_fema), the average
hourly relative humidity the week of collection
(Avg_HourlyRelativeHumidity_current), the average hourly wind speed the
week of collection (Avg_HourlyWindSpeed_current), the average number of
daylight hours per day (average_daylight_hours), the average number of
MEL captured by trap (Average_MEL), and the proportion of MEL captured
out of the total number of mosquitoes trapped (MEL_Prop).

@@

![](images/rmse_table-03.png){fig-align="center"}

@@

The Root Mean Square Error (RMSE) comparison outlined in the table
provides a clear perspective on the efficacy of different predictive
models in the context of Eastern Equine Encephalitis (EEE) surveillance
in southeastern Massachusetts. The null model, which served as a
baseline with an RMSE of 2.40, indicates the level of error absent any
predictive influence. This baseline was pivotal for evaluating
enhancements brought by more complex models.

The Generalized Additive Model (GAM), which incorporated eight predictors selected based on their statistical significance from preliminary analyses, demonstrated concerning signs of overfitting. While the model achieved a low RMSE of 1.7 on the training data, indicating a good fit to this dataset, its performance on the testing data was substantially worse, with an RMSE of 3.44. This significant increase in RMSE when applied to the testing data, compared to the training data, suggests that the model, despite fitting the training data well, fails to generalize to new, unseen data. This discrepancy in performance is further highlighted by the testing RMSE being higher than the RMSE of the null model at 2.40, underscoring the limitations of the GAM in this scenario.

Subsequent models such as the Random Forest, LASSO, Generalized Additive
Model with Sparsity (GAMS), and Gradient Boosting Machine (GBM) were
tested to potentially mitigate overfitting and enhance prediction
accuracy. Notably, the Gradient Boosting Machine model achieved the
lowest testing RMSE of 2.14, suggesting a model with better
generalizability compared to others.

Given that the average MIR value is 1.6, an RMSE of over 2 suggests that the models are not accurately capturing the variability in the data and are making predictions that are significantly different from the observed values.

Each model's performance underlines the complex interplay of various
ecological and environmental factors in predicting EEE transmission
cycles.

@@

## Conclusions

While all models
demonstrated improvement over null model, none of them were robust
enough to be used for reliable prediction. In the next phases of this
project, the data will be expanded to include more years and cover a
larger portion of Massachusetts. In additon to increasing the the number
of observations, other sources of data will be incorporated. Including
virus activity from surrounding regions and incorporating bird population data in to the models could provide more insight into the relationship between the predictors and virus activity.
